Lessons learned from bugs in models of human history

### 2020-08-05 paper.tex
Minor tweaks.

\title{Lessons learned from bugs in models of human history}
\author[1]{Aaron P. Ragsdale}
\author[1]{Dominic Nelson}
\author[1,$\star$]{Simon Gravel}
\author[2,$\star$]{Jerome Kelleher}
\affil[1]{McGill Genome Centre and Department of Human Genetics,
McGill University, Montr\'{e}al, Qu\'{e}bec, H2X 3C9, Canada}
\affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, Oxford, \aprcomment{postal code} United Kingdom}
\affil[$\star$]{Joint senior authors, listed alphabetically}
\maketitle

@@

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, such as inferring gene genealogies from sequencing data~\citep{kelleher2019inferring},
determining the age of a genetic variant~\citep{albers2020dating}, 
or studying the power of variant association tests~\citep{tong2020population},
and the model was used as an example of a complex population history.
\cite{zhou2018popdemog} used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
Finally, \cite{pfaffelhuber2020choose}
used simulations of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

%\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
\begin{Verbatim}[xleftmargin=.5in]
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{Verbatim}
%\end{minted}
Note that the migration rate change on 15 was missing in the incorrect model.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
%\begin{minted}[highlightlines={8}]{python}
\begin{Verbatim}[xleftmargin=.5in]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{Verbatim}
%\end{minted}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten as
%\begin{minted}[]{python}
\begin{Verbatim}[xleftmargin=.5in]
demography = msprime.Demography(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(demography=demography)
\end{Verbatim}
%\end{minted}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
library, which removes the need to re-implement demographic models from scratch.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
However, \stdpopsim has QC-verified models available for only a handful of published
models and species, so it might not meet the needs of a particular simulation study.
If the required model is not present in \stdpopsim, then extra care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis scale much
easier for large sample sizes~\citep{ralph2020efficiently}, so that $F_{ST}$,
heterozygosity, the allele frequency spectrum, or the distribution of coalescence
times can be rapidly computed. However, the problems would have been
apparent if pilot simulations with smaller sample sizes were undertaken and analysed,
as simulations and analysis of small samples sizes is very efficient.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.

>>>>

\title{Lessons learned from bugs in models of human history}
\author[1]{Aaron P. Ragsdale}
\author[1]{Dominic Nelson}
\author[1,$\star$]{Simon Gravel}
\author[2,$\star$]{Jerome Kelleher}
\affil[1]{McGill Genome Centre and Department of Human Genetics,
McGill University, Montr\'{e}al, Qu\'{e}bec, H2X 3C9, Canada}
\affil[2]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
University of Oxford, Oxford, OX3 7LF, United Kingdom}
\affil[$\star$]{Joint senior authors, listed alphabetically}
\maketitle

@@

Even though the error has a limited effect on simulated data, the tutorial code has been
copied many times and used in publications. By searching for some identifying strings
from the model definition on GitHub,
we found 32 repositories containing either direct copies of the
erroneous model code, or code that was obviously derived from it.
(We have opened issues on each of these repositories to alert the authors.)
In most cases the publications used simulations to test a non-demographic inference
method, such as inferring gene genealogies from sequencing data~\citep{kelleher2019inferring},
determining the age of a genetic variant~\citep{albers2020dating},
or studying the power of variant association tests~\citep{tong2020population},
and the model was used as an example of a complex population history.
\citet{zhou2018popdemog}~used the incorrect model as an example
of how their method for visualising demographic models can support
\msprime\ input.
Finally, \citet{pfaffelhuber2020choose}~used simulations
of the incorrect model demography to evaluate
their method for choosing ancestry informative markers.
Given the very subtle effect of the incorrect
model on demography (and the fact the method was evaluated using other
simulations and real data), it seems unlikely that the model details
had any qualitative effect on their conclusions.

@@

%\begin{minted}[linenos,highlightlines={15},numbersep=5pt]{python}
\begin{Verbatim}[xleftmargin=.5in]
demographic_events = [
    # CEU and CHB merge into B with rate changes at T_EU_AS
    msprime.MassMigration(
        time=T_EU_AS, source=2, destination=1, proportion=1.0),
    msprime.MigrationRateChange(time=T_EU_AS, rate=0),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(0, 1)),
    msprime.MigrationRateChange(
        time=T_EU_AS, rate=m_AF_B, matrix_index=(1, 0)),
    msprime.PopulationParametersChange(
        time=T_EU_AS, initial_size=N_B, growth_rate=0, population_id=1),
    # Population B merges into YRI at T_B
    msprime.MassMigration(
        time=T_B, source=1, destination=0, proportion=1.0),
    msprime.MigrationRateChange(time=T_B, rate=0), # NB This line was missing!!!
    # Size changes to N_A at T_AF
    msprime.PopulationParametersChange(
        time=T_AF, initial_size=N_A, population_id=0)
]
\end{Verbatim}
%\end{minted}
Note that the migration rate change on line 15 was missing in the incorrect model.
The release of \msprime\ version~1.0 will introduce a \texttt{PopulationSplit} event,
which more intuitively links the movement of lineages with appropriate changes in
migration rates at the time of the split.

@@

From a software perspective, this error was easy to make
and could have been prevented by better API design.
The original \msprime\ API requires the user to pass
three separate parameters to specify a demographic model. For example,
%\begin{minted}[highlightlines={8}]{python}
\begin{Verbatim}[xleftmargin=.5in]
dbg = msprime.DemographyDebugger(
  population_configurations=population_configurations,
  demographic_events=demographic_events,
  migration_matrix=migration_matrix)
dbg.print_history()
ts = msprime.simulate(
  population_configurations=population_configurations,
  demographic_events=demographic_events, # Missing in Martin et al. 2017
  migration_matrix=migration_matrix)
\end{Verbatim}
%\end{minted}
Note that the same three parameters must be passed to both the debugger
and the simulate function.
To help prevent such errors,
\msprime~ver.~1.0 will introduce a Demography class. The above
snippet would then be rewritten (approximately) as
%\begin{minted}[]{python}
\begin{Verbatim}[xleftmargin=.5in]
demography = msprime.Demography(
  populations=populations,
  events=demographic_events,
  migration_matrix=migration_matrix)
dbg = demography.debug()
dbg.print_history()
ts = msprime.simulate(
  # other parameters defining samples etc
  demography=demography)
\end{Verbatim}
%\end{minted}
This simple change to the interface makes it much less likely
that different models are passed to the debugger and simulator,
reducing the potential for error.

@@

Firstly, if possible, we recommend using the \stdpopsim~\citep{adrion2019community}
library, which removes the need to re-implement demographic models from scratch.
This is a ``standard library" of quality-controlled
models and simulation resources for a growing number of commonly studied species.
The resource is built around an open-source community development model
with rigorous code review and quality control procedures.
For example, when a contributor adds a new model it is not fully integrated
into the catalog until a second, entirely independent, implementation of the
model is provided by another developer. If the original and ``QC" versions
of the model are precisely equal, then we can have a reasonable degree
of confidence in correctness. It is through this QC process that we discovered
the misspecified model in \msprime's documentation.
However, \stdpopsim\ has QC-verified models available for only a handful of published
models and species, so it might not meet the needs of a particular simulation study.
If the required model is not present in \stdpopsim, then extra care should
be taken to validate the implementation.
This could include verification through code review or comparison with an
independent implementation of the model. If the model is of wider interest to
the community, we would encourage developers to contribute it to \stdpopsim:
at the very least, the quality-control procedures in place provide a strong
reassurance of correctness.

Secondly, regardless of whether the model has been implemented
locally or obtained from a resource like \stdpopsim,
basic statistical validation should \emph{always} be performed on the
results. Errors are all too easy to make, and an analysis of the
basic statistical properties of the simulations is essential
due diligence. For example, a brief analysis of $F_{ST}$ values
would likely would have prevented the error in~\citet{martin2017human}.
It is understandable that such analyses were not undertaken in this
case, as processing such a huge dataset (600,000 samples) was
highly challenging. Recent progress has made statistical analysis at this scale much
easier~\citep{ralph2020efficiently}, so that $F_{ST}$,
heterozygosity, the allele frequency spectrum, or the distribution of coalescence
times can be rapidly computed. However, the problems would have been
apparent if pilot simulations with smaller sample sizes were undertaken and analysed,
as simulations and analysis of small samples sizes is very efficient.
Graphical inspection of demographic models can also help identify issues, especially if
visualization can be automated. The demography debugging tool in \msprime\
summarizes demographic events occurring over time, and various efforts are underway to facilitate
the specification and visualization of population genetics models, such as~\citet{zhou2018popdemog}
and the \texttt{demography} package used in this paper.
